    
    
    

    
    This dissertation would not have been possible without the indefatigable
support of so many people. I would like to first and foremost thank
Panos Mavromatis for providing such a steadfast and invaluable source of
mentorship, encouragement, and guidance during the course of this entire
project through its many stages and developments.~

I thank my other committee members, Sarah Marlowe, Mary Farbood, and
Laurence Maloney for the many hours of time and outpouring of knowledge,
as well as their discerning input and careful criticism. I am grateful
to Eero Simoncelli for his generosity, and for providing me with so many
resources and opportunities. I also owe a huge portion of
gratitude~towards Marilyn Nonken for her expert guidance, and Eduardus
Halim for his unparalleled dedication and belief in me.

Jonathan Winters, Paul Fan, Marshall Krinitz, as well as the rest of the
NYU Psych IT team have been an incredible source of support, and have
constantly gone beyond the call of duty to selflessly help researchers
and students. A special thanks to Richie Berger and Mark Barnes as well
for being a most fantastic night shift team, and for their emotional
encouragement and camaraderie. I would like to also thank Lennart Belok
for his kindness and care towards me, especially during my injury
recovery phase.

I especially thank~Tanubot and Rodentia and all the tanukis and rodents
of the world. Lastly, I thank my mother, Chi-Hsiang Lin, and father,
Ren-Jang Lin, for their years of unconditional~and unwavering love and
devotion.



    % Add a bibliography block to the postdoc
    
    
    
